---
title: "RGPR tutorial - RGPRsurvey, adding coordinates & processing"
author: "Emanuel Huber (emanuel.huber@alumni.ethz.ch)"
date: '`r format(Sys.Date(), "%d %B %Y")`'
output:
  html_document:
    fig_caption: yes
    includes:
      in_header: header.html
    keep_md: yes
    number_sections: yes
    self_contained: no
    theme: default
    toc: yes
  pdf_document:
    highlight: tango
    number_sections: yes
    toc: yes
linkcolor: red
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/media/huber/Elements/UNIBAS/software/codeR/package_RGPR/RGPR-gh-pages_tools/2014_04_25_frenke") 
```

***

**Note**: 

* This R-package is still in development, and therefore some of the functions may change in a near future. 
* If you have any questions, comments or suggestions, feel free to contact me (in english, french or german): <emanuel.huber@alumni.ethz.ch>.

# Objectives of this tutorial
* Learn how to add coordinates to the GPR data.
* Learn how to plot the GPR data from your survey all together.

However, this tutorial will not explain you the math/algorithms behind the different processing methods.



# Preliminary

* Download the data [2014_04_25_frenke.zip](http://emanuelhuber.github.io/RGPR/2014_04_25_frenke.zip)
* Unzip the data

## File organisation
I suggest to organise your files and directories as follows:
```
/2014_04_25_frenke
    /processing      (here you will save the processed GPR files)
    /rawGPR          (the raw GPR data, never modify them!)
    /coord           (coordinates data)
        /FID         (fiducial marker files)
        /FIDmod      (modified fiducial marker files)
        /shapefiles  (here you will export the coordinate shapefiles)
        /topo        (ASCII files containing the trace coordinates)
    measured_coordinates.txt  (the topographic field measurements)
    RGPR_tutorial.R           (this is you R script for this tutorial)
```


## Install/load the necessary packages
Load the packages `RGPR` and `rChoiceDialogs` (`rChoiceDialogs` provides a collection of portable choice dialog widgets):
```{r, warning=FALSE, message=FALSE}
library(RGPR)   # load RGPR in the current R session
library(rChoiceDialogs)
```

[optionally] If `RGPR` is not installed, follow the instructions of the tutorial "Getting started" to install it.
[optionally] If R answers you `there is no package called 'rChoiceDialogs'` you need first to install `rChoiceDialogs`, either through your R software or directly in R with:
```{r, eval=FALSE}
install.packages("rChoiceDialogs")
```

The warnings that R shows can be ignored.

## Set the working directory
The working directory must be correctly set to use relative filepath. The working directory can be set either in your R-software or in R directly with (of course you need to change the filepath shown below):
```{r, eval=FALSE}
myDir <- file.path(c("/media/huber/Elements/UNIBAS/software/codeR",
                     "package_RGPR/RGPR-gh-pages/2014_04_25_frenke"))
setwd(myDir)    # set the working directory
getwd()         # Return the current working directory (just to check)
```

[optionally] Alternatively, you can use an interactive dialog box from the R-package `rChoiceDialogs`:
```{r, eval=FALSE}
myDir <- rchoose.dir(default = "/home/huber/WORK/UNIBAS/RESEARCH/RGPR/")
setwd(myDir)    # set the working directory
getwd()         # Return the current working directory (just to check)
```

# Read all the GPR data in a object of the class `GPRsurvey`
An object of the class `GPRsurvey` is like an index that contains some of the meta-data of several GPR data recorded during one survey. With the class `GPRsurvey` you have an overview of all your data, you can compute the positions of the profile intersections, plot a top view of the survey and plot the data in 3D with open-GL (implemented in the R-package `RGL`).

## Set the filepaths
Read all the GPR records (".DT1") located in the directory 
`/rawGPR` with the exception of the file `CMP.DT1` and create an object of the 
class `GPRsurvey`. To indicate the filepaths of the GPR data to R you have the
following options:

* use a dialogue window

```{r, echo=TRUE, eval=FALSE}
# select all the GPR files except CMP.DT1
LINES <- rchoose.files(caption = " DT1 files",filters = c("dt1","*.dt1"))
```

* write manually all the filepath in a list

```{r, echo=TRUE}
# select all the GPR files except CMP.DT1
LINES <- c()    # initialisation
LINES[1] <- file.path(getwd(), "rawGPR/LINE00.DT1")
LINES[2] <- file.path(getwd(), "rawGPR/LINE01.DT1")
LINES[3] <- file.path(getwd(), "rawGPR/LINE02.DT1")
LINES[4] <- file.path(getwd(), "rawGPR/LINE03.DT1")
LINES[5] <- file.path(getwd(), "rawGPR/LINE04.DT1")
```

* extract automatically all the GPR file from the directory

```{r, echo=TRUE}
# list of the filepath
allFilesinDir <- list.files(file.path(getwd(), "rawGPR"))
allFilesinDir
# now, select only the file ending with .DT1 and without "CMP" 
# in their names
selDT1 <- grepl("(.DT1)$", allFilesinDir, ignore.case = TRUE) & 
          !grepl("CMP", allFilesinDir, ignore.case = TRUE)
LINES <- file.path(getwd(), "rawGPR", allFilesinDir[selDT1])
```

## Create an object of the class `GPRsurvey`
```{r, echo=TRUE}
mySurvey <- GPRsurvey(LINES)
```

Have a look at the newly created object:
```{r,echo=TRUE}
mySurvey
```

You can see that no coordinates (x,y,z) are associated with the GPR data. 
Therefore, if you try to plot the suvey you will get:
```{r, eval = FALSE, echo=TRUE, fig.cap = "plot survey"}
plot(mySurvey, asp=1) # throw an error
```

Note that the object `mySurvey` only contains the meta-data and a link
to the GPR files (that are stored in your working directory).
But `mySurvey` does not contains the GPR data itself (i.e. the traces). 
However, we can ask `mySurvey` to read the data and return it in the form
of an object of the class `GPR`. There are two possibilities:

* Either subset `mySurvey` with `[[ ]]`

```{r, echo=TRUE}
A02 <- mySurvey[[3]]
A02
```

* Or get the `GPR` object with the function `getGPR()`

```{r, echo=TRUE}
A02 <- getGPR(mySurvey, id = "LINE02")
A02
```

You can also directly plot the GPR data with:
```{r, echo=TRUE}
# instead of 'A02 <- mySurvey[[3]]' and 'plot(A02)' do:
plot(mySurvey[[2]]) 
```

# Add coordinates
There are two options: (1) either you already have the coordinates of each 
traces or (2) you have the coordinates of points that were marked in
the GPR data using fiducial marker.

## Option 1: all the trace coordinates are already known

We assume that for each GPR record there is a file containing the (x, y, z) 
coordinates of every traces. The header of these files is "E", "N", "Z" 
instead of "x", "y", "z" because in topography "x" sometimes designates the 
North ("N") and not the East ("E") as we would expect. The designation
"E", "N", "Z" is less prone to confusion and therefore we chose it!

1. Define the filepaths to the topo files:
```{r, echo=TRUE}
# select all the GPR files except CMP.DT1
TOPO <- c()    # initialisation
TOPO[1] <- file.path(getwd(), "coord/topo/LINE00.txt")
TOPO[2] <- file.path(getwd(), "coord/topo/LINE01.txt")
TOPO[3] <- file.path(getwd(), "coord/topo/LINE02.txt")
TOPO[4] <- file.path(getwd(), "coord/topo/LINE03.txt")
TOPO[5] <- file.path(getwd(), "coord/topo/LINE04.txt")
```

2. Read all the files with the funciton `readTopo()` that creates a list whose 
elements correspond to the GPR record and contain all the trace coordinates:
```{r, echo=TRUE}
TOPOList <- readTopo(TOPO, sep = "\t")
```

3. Set the list of coordinates as the new coordinates to the GPRsurvey object:
```{r, echo=TRUE, warning=FALSE, message=FALSE, results = 'hide'}
coords(mySurvey) <- TOPOList
```

## Option 2: some trace coordinates exist
The file `coord/measured_coordinates.txt` shows some of the surveyed
coordinates. The content of the file is shown below.

```
GPR survey 25 April 2014, Frenkental
Coordinates reference system: CH1903+ / LV95

XLINE00
START 2622172.58, 1256908.26 346.7 
FID1  2622218.98, 1256906.46 345.9
END   2622229.68, 1256905.16 345.9

XLINE01
START 2622233.08, 1256905.76 	346 
END 	2622244.28, 1256905.16 346

XLINE02
START 2622229.98, 1256905.56 345.9
END   2622226.08, 1256842.96 346.7

XLINE03
START   2622226.08, 1256842.96 346.7
FID1	2622265.48, 1256843.26 344
END		2622269.08, 1256842.96 343.4 

XLINE04
START 	2622262.98, 1256834.06 343.8
FID2 	2622265.48, 1256843.26 344
END  	2622300.41, 1256921.53 343.5 
```
You observe that the coordinates of the begining and end of each GPR profile
are known and that the coordinates of some fiducial markers were also
surveyed. 

1. Now, you have to indicate R to which traces the coordinates correspond.
That means, for each GPR data you have to create a file that contains the
coordinates with the associated trace number. This is a lot of work.
To help you to create these files, use the function `exportFid()`. 
This function create for each GPR data a file containing the trace number, 
position for the start and end positions
as well as for the fiducial makers.

```{r, echo=TRUE, warning=FALSE, message=FALSE}
exportFid(mySurvey, fPath = file.path(getwd(), "coord/FID/"))
```

2. Now, add to each FID files three columns corresponding to the trace 
coordinates. These columns must have as header "N", "E" and "Z" (for North,
East and vertical elevation). Why not "x", "y" and "z"? Because in topography,
"x" defines sometimes the "North", so it may be confusing...

Here is the FID for the GPR data `LINE04` (`coord/FID/LINE04.txt`):

```
TRACE,POSITION,COMMENT
1,0,START
91,22.5,F1
100,24.75,F2
122,30.25,F3
445,111,END
```

The same file with the three columns added:
```
TRACE,POSITION,COMMENT,E,N,Z
1,0,START,2622262.98,1256834.06,343.8
100,24.75,F2,2622265.48,1256843.26,344
445,111,END,2622300.41,1256921.53,343.5 
```

Note that the two lines with the fiducial markers F1 and F3 were removed
as no coordinates are available for these markers. Save the modified files
in the directory `coord/FIDmod`.

3. Read the modified fiducial marker files usind the funtion `readFID()`:

```{r, echo=TRUE}
FidFiles <- c()    # initialisation
FidFiles[1] <- file.path(getwd(), "coord/FIDmod/LINE00.txt")
FidFiles[2] <- file.path(getwd(), "coord/FIDmod/LINE01.txt")
FidFiles[3] <- file.path(getwd(), "coord/FIDmod/LINE02.txt")
FidFiles[4] <- file.path(getwd(), "coord/FIDmod/LINE03.txt")
FidFiles[5] <- file.path(getwd(), "coord/FIDmod/LINE04.txt")
```

```{r, echo=TRUE, warning=FALSE, message=FALSE, results = 'hide'}
FIDs <- readFID(FidFiles)
```

If R throw the following error message 
"Error in readFID(FidFiles) : The headers should be "E","N","Z","TRACE"!"
check that the three columns have been added to all the files and that
the headers of these columns are correctly set ("N", "E", "Z"; the order
does not matter).

4. Interpolate the coordinates of the traces for all the GPR profiles 
according to the modified fiducial marker files. The function `interpPos()`
interpolate the position of the traces from the known trace positions and
add the interpolated trace position to the object `mySurvey`. 
```{r, echo=TRUE}
# interpolating the positions of the traces between the fiducials for each
# GPR-lines and adding the position to the survey.
# + compute the intersection between the GPR-lines
# windows open for checking purposes
# dx should be between 0.1 m and 0.5 m
mySurvey <- interpPos(mySurvey, FIDs) 
```

The function `interpPos()` prints for every GPR record the mean trace spacing
as well as the trace spacing range. Normally, these values should be close
to the operating settings. In this case, the trace spacing was set
equal to $0.25\,m$ on the field. The trace spacing values for `XLINE00`,
`XLINE01` and `XLINE02` looks good. However, the trace spacing for `XLINE04`
and more particularly for `XLINE03` could be critic (the smallest trace spacing
values are very low). You should check and if necessary correct the topographic
data...

## Setting the coordinate reference system

Setting the coordinate reference system is important when exporting the 
coordinate data in geospatial data format, because it allows the coordinates
to be correctly projected in other coordinate reference systems.
The topographic data were measured within the new Swiss coordinate system 
(datum: CH1903+, reference frame: LV95) that can be defined with the
code EPSG $2056$ that corresponds to the new Swiss coordinate system.

```{r, echo=TRUE}
crs(mySurvey) <- "+init=epsg:2056"
```

## Export the coordinates
To export the coordinates as shapefiles (one shapefile for all the GPR 
records), enter:
```{r, echo=TRUE}
exportCoord(mySurvey, fPath="coord/shapefiles/frenke")
```


To export the coordinates as ASCII (`.txt`) files (on file per GPR record):
```{r, eval = FALSE, echo=TRUE}
exportCoord(mySurvey, folder="coord/topo", type="ASCII")
```


## Remark
Note that the coordinates are added to the object `mySurvey` but not to the 
GPR file. Unless you save the GPR data you will lose the coordinates when
you will quit R. To save the GPR data, see [Save, export][].

# Plot functions

## Plot the survey
Use the `plot()` function
```{r, echo=TRUE, warning=FALSE, message=FALSE, results = 'hide', fig.cap = "plot survey"}
plot(mySurvey)
```
The red arrows indicate the direction of the survey, the red dots the fiducial markers and the circles the GPR profile intersections.

## Plot the GPR data

To plot the first GPR record, enter:
```{r, echo=TRUE, fig.cap = "plot GPR data"}
plot(mySurvey[[1]], addTopo=TRUE)
```

## Three-dimensional plot of the GPR data
To plot all the GPR records with the topographic information in 3D, enter:
```{r, echo=TRUE, eval = FALSE}
plot3DRGL(mySurvey, addTopo = TRUE)
```

Enlarge the window, use the mouse to move the view and zoom in.


## <a name="pookie"></a>Processing, save and PDF export
Once you found a satisfactory combination of processing steps, you can apply
them all the GPR data. Here is an example

Create a sub-directory in the `/processing` directory (name it `mySurveyProc`):

```{r, echo=TRUE, eval = TRUE, warning=FALSE, message=FALSE, results = 'hide'}
procDir <- file.path(getwd(), "processing/mySurveyProc/")
dir.create(file.path(procDir),showWarnings = TRUE)
```

Now apply the processing step with a loop to all GPR data indexed by `mySurvey`:

```{r, echo=TRUE, eval = FALSE, results = 'hide'}
for(i in seq_along(mySurvey)){
  A <- mySurvey[[i]]            # get GPR-line no. i
  cat("processing of line", name(A),"      ")
  A <- dcshift(A, 1:100)    	  # DC-shift
  A <- gain(A, type = "power", alpha = 1, te = 200, tcst = 100)
  A <- fFilter(A, f= c(150, 200), type = "low", plotSpec = FALSE)
  A <- gain(A, type = "exp", alpha = 0.1, t0 = 50, te = 180)
  A <- dewow(A, type = "MAD", w = 50)              # dewow
  A <- filter2D(A, type = "median3x3")
  # export PDF
 exportPDF(A, clip = 30, fPath = file.path(procDir, name(A)), addTopo = TRUE,
          lwd = 0.5, ws = 1.5)
  # save the processed GPR-line into ".rds" format
  writeGPR(A, fPath = file.path(procDir, paste0(name(A),".rds")), 
         format = "rds", overwrite = TRUE)
  cat("!\n")
}
``` 


## Read again
Next time you can directly load the processed files as follows:
```{r, echo=TRUE}
# select all the GPR files except CMP.DT1
procLINES <- c()    # initialisation
procLINES[1] <- file.path(procDir, "LINE00.rds")
procLINES[2] <- file.path(procDir, "LINE01.rds")
procLINES[3] <- file.path(procDir, "LINE02.rds")
procLINES[4] <- file.path(procDir, "LINE03.rds")
procLINES[5] <- file.path(procDir, "LINE04.rds")
```

```{r, echo=TRUE, eval = FALSE}
procSurvey <- GPRsurvey(procLINES)
```


and check the results:

```{r, echo=TRUE, eval = FALSE}
plot(procSurvey[[1]], addTopo=TRUE)
plot3DRGL(procSurvey, addTopo = TRUE)
```
